---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bis620.2023

<!-- badges: start -->
[![R-CMD-check](https://github.com/leafsphere/bis620.2023_assn3/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/leafsphere/bis620.2023_assn3/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `bis620.2023` is to provide a user-friendly interface that allows the viewer to explore a large database of clinical studies. Inside lies an R Shiny app that serves as an interactive dashboard, allowing the user to query trials based on their keyword(s) of interest and subsequently shows relevant data visualizations and tables pertaining to the results, such as what phases the studies were in, what countries were involved, and what conditions were examined. 


## Installation

You can install the development version of bis620.2023_assn3 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("leafsphere/bis620.2023_assn3")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(bis620.2023)
accel |> 
  head(100) |> 
  plot_accel()
```

